---
title: "Musical Genres"
author: "Neetu Regmi, Cassie Tangen, Oliver Titus"
date: "10/20/2020"
output:
  beamer_presentation: default
  ioslides_presentation: default
subtitle: UNL STAT 850
---

## Introduction

- Music

>let's insert a picture here of spotify's logo?


## Data collection

- TidyTuesday
- Spotifyr


## Genres

- Rock
- R&B
- Rap
- EDM
- Latin
- Pop

## Variables

- Duration
- Danceability
- Energy
- Loudness
- Speechiness
- Popularity

## Proposed Study

- Determine if there are defining characteristics for each genre.
- Determine if a relationship between track traits and track popularity exists.

## Boxplot

```{r,include=FALSE,echo=FALSE}
# load the data
spotify_songs <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')
```

```{r, warning=F, message=F, echo=F}
library(tidyverse)
library(dplyr)
library(knitr)
library(broom)
library(moderndive)
ggplot(spotify_songs, aes(y=duration_ms, fill=playlist_genre)) + geom_boxplot() + ggtitle("Duration for each Genre") + ylab("Milliseconds (ms)") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```


## Boxplot

```{r, warning=F, message=F, echo=F}
ggplot(spotify_songs, aes(y=loudness, fill=playlist_genre)) + geom_boxplot() + ggtitle("Loudness for each Genre") + ylab("Decibals (dB)") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```


## Boxplot 

```{r, warning=F, message=F, echo=F}
ggplot(spotify_songs, aes(y=danceability, fill=playlist_genre)) + ggtitle("Danceability for each Genre") + geom_boxplot() + ylab("Danceability") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```


## Boxplot 

```{r, warning=F, message=F, echo=F}
ggplot(spotify_songs, aes(y=energy, fill=playlist_genre)) + 
    geom_boxplot() + ggtitle("Energy for each Genre") + ylab("Energy") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```


## Boxplot


```{r, warning=F, message=F, echo=F}
ggplot(spotify_songs, aes(y=speechiness, fill=playlist_genre)) + 
    geom_boxplot() + ggtitle("Speechiness for each Genre") + ylab("Speechiness") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```


## Boxplot

```{r, warning=F, message=F, echo=F}
ggplot(spotify_songs, aes(y=track_popularity, fill=playlist_genre)) + geom_boxplot() + ggtitle("Popularity for each Genre") + ylab("Track Popularity") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```

## Summary Statistics


```{r, warning=F, message=F, echo=F}
summary_table1<- group_by(spotify_songs, playlist_genre) %>% summarise(count=n(), mean_loudness=mean(loudness), mean_duration_in_ms=mean(duration_ms), mean_danceability=mean(danceability))

summary_table2<- group_by(spotify_songs, playlist_genre) %>% summarise(count=n(), mean_speechiness=mean(speechiness), mean_energry=mean(energy), mean_mean_popularity=mean(track_popularity))

summary_table1 %>% kable()

```

## Summary Statistics


```{r, warning=F, message=F, echo=F}

summary_table2 %>% kable()
```

## ANOVA Test Results

- Performed ANOVA test to determine if there is a statistical difference between genres across all track trait variables.
-Based on F-test, We were able to reject the null Hypothesis i.e mean of all variables are same.
-There is enough evidence to support that there is a difference with all of these characteristics across each genre.

## Linear Regression
```{r, echo=FALSE, message=FALSE}
ggplot(spotify_songs, aes(x=danceability, y=track_popularity)) + 
    geom_point()+geom_smooth(method=lm) + ggtitle("Track Popularity vs. Danceability") + xlab("Danceability") + ylab("Track Popularity")

```

## Linear Regression
```{r, echo=FALSE, message=FALSE}
ggplot(spotify_songs, aes(x=energy, y=track_popularity)) + 
    geom_point()+geom_smooth(method=lm) + ggtitle("Track Popularity vs. Energy") + xlab("Energy") + ylab("Track Popularity")

```


## Linear Regression
```{r, echo=FALSE, message=FALSE}
ggplot(spotify_songs, aes(x=speechiness, y=track_popularity)) + 
    geom_point()+geom_smooth(method=lm) + ggtitle("Track Popularity vs. Speechiness") + xlab("Speechiness") + ylab("Track Popularity")
```


## Linear Regression

```{r, echo=FALSE, message=FALSE}
ggplot(spotify_songs, aes(x=duration_ms, y=track_popularity)) + 
    geom_point()+geom_smooth(method=lm) + ggtitle("Track Popularity vs. Duration") + xlab("Milliseconds (ms)") + ylab("Track Popularity")
```


## Linear Regresssion
```{r, echo=FALSE, message=FALSE}
ggplot(spotify_songs, aes(x=loudness, y=track_popularity)) + 
    geom_point()+geom_smooth(method=lm) + ggtitle("Track Popularity vs. Loudness") + xlab("Decibals (dB)") + ylab("Track Popularity")
```


## Conclusions

-From the study, we conclude that there is not one defining characteristics that will make track more popular.
- If we have more time, we would compare genres that are most similar to determine if the characteristics are same or different.
-Furthermore, it would be interesting to study other genres including some of the subgenres which will encompass more genre than in our study.
-In addition, it would be interesting to determine if there exists a relationship between the characteristics that are similar between genres.


## References

(n.d.). Retrieved from R Wrapper for the 'Spotify' Web API: https://www.rcharlie.com/spotifyr/
Mock, T. (2020, January 21). Retrieved from TidyTuesday: https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-01-21)
Spotify. (2020, November 10). Retrieved from Wikipedia: https://en.wikipedia.org/wiki/Spotify

