---
title: "Stat 850 Project Description"
author: "Neetu Regmi, Cassie Tangen, Oliver Titus ..."
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 3)
```

```{r load-sas-libraries, echo = F, message = F, warning = F}

saspath <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en" # Your SAS path goes here
sasopts <- "-ls 75"
knitr::opts_chunk$set(engine.path=list(sas=saspath, saslog=saspath, sashtml=saspath),
                      engine.opts=list(sas=sasopts, saslog=sasopts, sashtml=sasopts), 
                      comment=NA)
library(SASmarkdown)
```

## Instructions

Each member of your team should modify this document in some way and push their modifications to the repository in a separate commit. This will ensure that you have set your repository up in a way that ensures all group members are working with the same repository. 

Note that you must compile your readme (this document) for it to properly display as part of your github repository. 

Once you have received feedback on your project proposal (via Canvas) you may alter this README so that it describes your final project instead of the project proposal. 

## Data Set

spotify_songs <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')

Data from TidyTuesday about songs on Spotify.
There are many variables included which are all described by TidyTuesday in the 01-21-2020 folder titled song genres.


## Potential Topics to Explore Using the Data Set
Describe avenues you might explore using the data

Exploring if there is difference in song length based on genre.

Explore the popularity of each genre.

Explore the loudness of different genre.

Explore track artist of different genre.

Explore a potential relationship between tempo and song duration.

Explore a potential relationship between tempo and danceability,

## Group Members
List all of the project contributors here.
Neetu Regmi 
Oliver Titus 
Cassie Tangen
