---
title: "Stat 850 Project Description"
author: "Neetu Regmi, Cassie Tangen, Oliver Titus ..."
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 3)
```

```{r load-sas-libraries, echo = F, message = F, warning = F}

saspath <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en" # Your SAS path goes here
sasopts <- "-ls 75"
knitr::opts_chunk$set(engine.path=list(sas=saspath, saslog=saspath, sashtml=saspath),
                      engine.opts=list(sas=sasopts, saslog=sasopts, sashtml=sasopts), 
                      comment=NA)
library(SASmarkdown)
```

## Instructions

Each member of your team should modify this document in some way and push their modifications to the repository in a separate commit. This will ensure that you have set your repository up in a way that ensures all group members are working with the same repository. 

Note that you must compile your readme (this document) for it to properly display as part of your github repository. 

Once you have received feedback on your project proposal (via Canvas) you may alter this README so that it describes your final project instead of the project proposal. 

## Data Set

https://raw.githubusercontent.com/stat850-unl/project-team-con/master/crops.csv

Data is from Food and Agriculture Organization of the United Nations(FAO). We selected Canada and the United States for countries to study. We are looking at 5 crops that are commonly produced in Nebraska: Wheat, Maize(corn), Soybeans, Sugar Beets, and Sorghum. We selected all three elements provided by FAO which include: Area Harvested, Yield, and Production Quantity. These are listed in the value column of the csv that we are working with. We are looking at 40 years worth of data. The units for yield are per hectare 

## Potential Topics to Explore Using the Data Set
Describe avenues you might explore using the data

## Group Members
List all of the project contributors here.

