---
title: "STAT 850 Final Project"
author: "Neetu Regmi, Cassie Tangen, Oliver Titus"
keywords: [music, spotify, dance]
output: 
  pagedown::jss_paged:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = T)
options(digits = 3)
```

```{r load-sas-libraries, echo = F, message = F, warning = F}

saspath <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en" # Your SAS path goes here
sasopts <- "-ls 75"
knitr::opts_chunk$set(engine.path=list(sas=saspath, saslog=saspath, sashtml=saspath),
                      engine.opts=list(sas=sasopts, saslog=sasopts, sashtml=sasopts), 
                      comment=NA)
library(SASmarkdown)
```


# Introduction

Music plays a role in everyone’s life regardless of age, race, gender, social class or any other defining characteristic. With the expansion of the World Wide Web numerous online music streaming platforms have become a popular source of listening to music. One such platform is Spotify. Spotify first launched in 2008 and has grown to offer more than 60 million songs. Spotify contains songs from a variety of genres and keeps track of the popularity of each track along with various other characteristics of each track. As a group we were interested in studying these characteristics to determine if there are unique characteristics that make each genre unique, or if some of the characteristics are shared across the genres. (Spotify, 2020)

We were able to find a nice set of data from TidyTuesday. The data was shared in the Jan 21, 2020 TidyTuesday folder. The data was originally obtained by making use of the spotifyr package.(Mock, 2020) The spotifyr package was created by Charlie Thompson, Josiah Parry, Donal Phipps, and Tom Wolff. The spotifyr package allows the bulk pulling of features tracks from Spotify’s Web API. Since we are novices at using new packages, we decided just using the data collected by TidyTuesday using spotifyr would be sufficient for our study. We do recognize the fact that there is a potential bias to the data we are analyzing because we are not aware of the search parameters used by TidyTuesday when collecting this data using spotifyr.(R Wrapper for the 'Spotify' Web API, n.d.)

```{r, echo=F, message = F, warning = F}
# load the data
spotify_songs <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')
```

# Data Description

The genres that data was collected on include rock, R&B, rap, EDM, Latin, and pop. Each of these genre’s likely illicit thoughts about certain characteristics. For example, rock and EDM I tend to think of loud, whereas R&B I tend to think is softer. Rap is often thought to be almost spoken whereas Latin is much more musical with the vocals. Latin or pop are often thought as a genre for dancing and rock is not a genre many people dance to. Thinking about how each of these genres are so vastly different we are going to look at 5 characteristics that we feel would be different across the genres. 

The characteristics that we determined would be best to study were the duration, loudness, speechiness, energy, and danceability. We will look at each of these along with popularity, first to determine if there is a unique characteristic for each genre and then if there is a relationship between each of these characteristics and their overall popularity. We also look at the popularity of each genre to help us determine if a particular characteristic about the music tends to make the song more popular.

We should begin with an explanation of our variables. Duration is the measure of the length of the song and is reported in milliseconds. Danceability is a value between 0.0 and 1.0 where 0.0 is the least danceable and 1.0 is the most danceable. The danceability is determined by an algorithm that looks at the beat strength, tempo, rhythm stability, and overall regularity. Energy is a value between 0.0 and 1.0 that represents the intensity and activity. Energetic tracks are often fast and noisy. The loudness is measured in decibels (dB) and is the average of the decibels across the entire track. Values will typically range between -60 and 0 dB. Speechiness is the measure of the spoken word present in a song. The closer the value is to 1 the more speech-like the track. Songs above 0.66 are likely entirely spoken words. (Mock, 2020)

# Data Summary
We will look at the interesting variables mentioned above for each of the genre to attempt to determine a defining characteristic or two about each genre. Let's first look at a nice stacked boxplot for each of the variables that we are interested in exploring along with a boxplot of the popularity of each genre. These might give us a visual idea of which variables are key characteristics for each genre.

## Visual Exploration of Data

```{r, warning=F, message=F, echo=F}
library(tidyverse)
library(dplyr)
library(knitr)
library(broom)
library(moderndive)
ggplot(spotify_songs, aes(y=duration_ms, fill=playlist_genre)) + geom_boxplot() + ggtitle("Duration for each Genre") + ylab("Milliseconds (ms)") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```

As we inspect the boxplot of duration for each genre we see that the duration of the tracks we see that Rock and R&B have the longest tracks. EDM seems to have the shortest duration. It is important though to note that the differences between the genres is pretty small so there is likely very little difference statistically. It is interesting to note that the data appears to be approximately bell shaped with very little skewing due to outliers.


```{r, warning=F, message=F, echo=F}
ggplot(spotify_songs, aes(y=loudness, fill=playlist_genre)) + geom_boxplot() + ggtitle("Loudness for each Genre") + ylab("Decibals (dB)") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```

When inspecting the average loudness of tracks for each genre we discover a bit more variation than we saw with the duration. We note here that EDM appears to be the loudest with R&B the least loud. There are low outliers for each genre, meaning that each genre does have a few songs that are relatively quiet. Latin and R&B are the two genres with the quietest tracks.


```{r, warning=F, message=F, echo=F}
ggplot(spotify_songs, aes(y=danceability, fill=playlist_genre)) + ggtitle("Danceability for each Genre") + geom_boxplot() + ylab("Danceability") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```

After a quick look at the boxplot for the danceability of each genre, it is clear that rock is the least danceable. Latin is fairly danceable and surprisingly so is rap. We see that there are low outliers for each genre and that rock has a few high outliers as well. This shows us that each genre has a few songs that score poorly on the danceability rating. Also that rock does have a few songs that are fairly danceable even though as a genre rock appears to be the least danceable.

```{r, warning=F, message=F, echo=F}
ggplot(spotify_songs, aes(y=energy, fill=playlist_genre)) + 
    geom_boxplot() + ggtitle("Energy for each Genre") + ylab("Energy") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```

The boxplot of energy for each of each genre doesn't show us anything too surprising. There are low outliers for all of the genres, which lets us know that every genre has a few songs that are low energy. Just knowing what our genres are it should be expected that EDM would be high energy and R&B would be low energy. Also not surprising rock appears to be a bit higher in energy than all but EDM.Interestingly,we can see that R&B is the least skewed due to it's outliers.

```{r, warning=F, message=F, echo=F}
ggplot(spotify_songs, aes(y=speechiness, fill=playlist_genre)) + 
    geom_boxplot() + ggtitle("Speechiness for each Genre") + ylab("Speechiness") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```

When it comes to thinking about speechiness, one would of course guess that rap would be the speechiest. The boxplot clearly depicts and significant difference in the speechiness between the genres with rap having a much higher speechiness rating than the other genres. It is also clear from the boxplot that every genre actually has a few high outliers meaning every genre does have a few songs that might feel more like what we often think of as rap. Also since we see that rap even has high outliers it shows us that while we do commonly think of rap as a very speechy genre of music quite a few of the tracks are not as high on the speechiness scale as you would expect. Though this does make sense because every track has to have some good musical qualities to it.

```{r, warning=F, message=F, echo=F}
ggplot(spotify_songs, aes(y=track_popularity, fill=playlist_genre)) + geom_boxplot() + ggtitle("Popularity for each Genre") + ylab("Track Popularity") + scale_fill_discrete(name = "Genre", labels = c("EDM", "Latin", "Pop", "R&B", "Rap", "Rock"))
```

As we conclude our graphical comparisons we wanted to take a look at the popularity of tracks based on genre. It is pretty clear from the boxplot that EDM is the least popular with pop being the most popular genre. We are able to see that most of the genres have a fairly bell shaped distribution that might tend toward a uniform distribution. There are always some tracks with low popularity regardless of the genre and some tracks that are highly popular. It is interesting to note while looking at this boxplot that while EDM has the lowest median it actually rock that has the smallest maximum value for track popularity. 

Now that we have taken a look at the these boxplots we can would guess that there would be a statistical difference in the average danceability, energy, and speechiness based on genres. While we expect that the duration and loudness of each genre would statistically be roughly the same even though we can pick out a few that appear to be on the upper or lower end of the spectrum it is not a huge difference.

## Summary Statistics

We will be performing ANOVA tests to determine if the genres have the same average value for each of the variables that we are interested in. The ANOVA test looks at the means whereas the boxplots summarize the data by using the five number summary which showcases the minimum, first quartile, median, third quartile, and maximum of our data. The five number summary values are more likely to be skewed by outliers as compared to the mean. In the appendix you will find a table with the mean values for each genre for each of the variables we studied. When looking through the means for each characteristic it can be noted that there is at least one genre that appears to be different than the rest. 

From looking at the means we can see a few characteristics that appear to make each genre stand apart. A summary of these differences can be useful in coming to a hypothesis of what traits make each genre unique. Also by looking at the genres popularity we may find a few clues as to which traits could give us an indication of the popularity of the track.

EDM tends to be loud with high energy. EDM is the least popular genre that we studied. R&B tends to have longer tracks and is quiet compared to other genres R&B also is low on energy and on the longer end for duration. R&B is the second least popular genre in our study. Rock is the third least popular genre in our study, but it is only slightly more popular than R&B. Rock often is quiet compared to the other genres, but has a notably longer track length. 

Rap comes next in popularity rankings, it is the fourth least popular, making it the third most popular. Rap typically is very speechy and has a high danceability score. Latin typically is shorter than other genres and has a high danceability rating. Latin is the second most popular genre. The most popular genre is pop, though it only beats out Latin by a small margin. Pop tracks tend to be on the shorter end for track length and has very little speechiness. 

From this summary we can see that Latin and pop music share a few traits and R&B and rock share similar traits. We also notice that each of those pairs have similar popularity rankings. This would lead us to hypothesis that there are certain characteristics that might influence the popularity of the track.


```{r, warning=F, message=F, echo=F}
#creating the data frames that we used later on to make the calculations easier.

#not sure we actually use this later on at this point, but I'm going to leave it here for now. If we don't need it someone can move it to the bottom until we finalize our project.

spotify_songs_edm<-filter(spotify_songs, playlist_genre=="edm")
spotify_songs_latin<-filter(spotify_songs, playlist_genre=="latin")
spotify_songs_pop<-filter(spotify_songs, playlist_genre=="pop")
spotify_songs_rb<-filter(spotify_songs, playlist_genre=="r&b")
spotify_songs_rap<-filter(spotify_songs, playlist_genre=="rap")
spotify_songs_rock<-filter(spotify_songs, playlist_genre=="rock")

```


# Results

We ran a few different tests on our data to determine if there is a difference in our genres based on the 5 track characteristics we selected, then to determine if any of those characteristics had an impact on the overall popularity of the track.

First an ANOVA test was performed to determine if there is a statistical difference between the genres when looking at each of the track trait variables. We performed an ANOVA test on each of the characteristic. The results from the ANOVA test can be found in the appendix. Interestingly even though we guessed that there may not be a statistical difference with the duration and the loudness of tracks  between the genres we found in all of the overall F-tests that we should reject the null hypothesis that the mean speechiness, energy, danceability, loudness, duration, and popularity are the same. This means we should conclude that there is evidence to support that there is a difference with all of these characteristics across each genre. Since we found that there is a difference naturally we should investigate to see which genres are the same and which are different. To make this comparison we have used the Tukey test to determine which genres are different that the others.

The Tukey test gave us some interesting results and some results that we kind of expected from our graphical analysis of the data. When looking at the speechiness of the genres we found that each genre is statistically different when it comes to the level of speechiness in the tracks. This is not terribly surprising from looking at the boxplots. When we looked at each of the other track characteristics we found at least a pair of genres that were similar.

We were able to determine that pop and latin are similar in energy, loudness, and duration. We also found that rap and latin have similar danceability ratings as well as being similar in duration. Seeing as pop and latin were similar in duration and latin and rap are similar in duration it was not surprising when it was found that rap and pop also have tracks of similar length. When looking at the popularity of genres we found that pop and latin are similar and that rock and rap are similar. Though the strength of the similarities between rock and rap for popularity ratings is not near as strong as the similarity between pop and latin.

After analyzing the differences between the genres with the ANOVA test we find that pop and latin are very similar in characteristics and also have very similar popularity ratings. This leads us further to guess that there will be a relationship between the track characteristics and the popularity rating.

## Testing for linear relationships between track characteristics and popularity. 
As we perform tests to determine if there are linear relationships between the track characteristics and the popularity of the track we also looked at the scatterplots. For many of the scatterplots we were able to see fairly clearly that there would not be a linear relationship between the characteristics and the popularity. 

### Danceability
```{r, message=FALSE}
ggplot(spotify_songs, aes(x=danceability, y=track_popularity)) + 
    geom_point()+geom_smooth(method=lm) + ggtitle("Track Popularity vs. Danceability") + xlab("Danceability") + ylab("Track Popularity")

```

From the scatterplot, there appears to be no significant relationship between danceability and track popularity. When we performed a linear regression test we found that both the correlation coefficient and the coefficient of determination are both close to zero. Thus,the variation observed in track popularity is not explained by danceability. 

### Energy
```{r, message=FALSE}
ggplot(spotify_songs, aes(x=energy, y=track_popularity)) + 
    geom_point()+geom_smooth(method=lm) + ggtitle("Track Popularity vs. Energy") + xlab("Energy") + ylab("Track Popularity")

```


From the scatterplot, there appears to be slightly negative relationship between track popularity and energy. When we performed a linear regression test we found that both the correlation coefficient and the coefficient of determination are both close to zero. Thus, but variation in track popularity is not explained by energy.

### Speechiness
```{r, message=FALSE}
ggplot(spotify_songs, aes(x=speechiness, y=track_popularity)) + 
    geom_point()+geom_smooth(method=lm) + ggtitle("Track Popularity vs. Speechiness") + xlab("Speechiness") + ylab("Track Popularity")
```


From the scatterplot, it appears that there is no significant relationship between track popularity and speechiness. When we performed a linear regression test we found that both the correlation coefficient and the coefficient of determination are both close to zero. Therefore, we can say that any variation in track popularity cannot be explained by speechiness.

### Duration

```{r, message=FALSE}
ggplot(spotify_songs, aes(x=duration_ms, y=track_popularity)) + 
    geom_point()+geom_smooth(method=lm) + ggtitle("Track Popularity vs. Duration") + xlab("Milliseconds (ms)") + ylab("Track Popularity")
```


This scatterplot shows that there may be a negative relationship between duration of song and track popularity. When we performed a linear regression test we found that both the correlation coefficient and the coefficient of determination are both close to zero. Yet again, we can say that any variation in track popularity cannot be explained by duration.

### Loudness
```{r, message=FALSE}
ggplot(spotify_songs, aes(x=loudness, y=track_popularity)) + 
    geom_point()+geom_smooth(method=lm) + ggtitle("Track Popularity vs. Loudness") + xlab("Decibals (dB)") + ylab("Track Popularity")
```


There appears to be no significant relationship between track popularity and loudness. When we performed a linear regression test we found that both the correlation coefficient and the coefficient of determination are both close to zero. So, most of variation is not explained by the loudness which suggests that there might other variables at play.

# Conclusions

From our study we are able to say that while there are a few defining characteristics for each genre those characteristics do not predict the popularity of the track. It was interesting though to note the unique characteristics that made each genre have it's own appeal. Ultimately it is not terribly surprising that there is not one characteristic that will make a song more popular. With the wide range of people who listen to music there should be a wide range of traits that are desirable and will result in a popular song.

If we had time to further our study we would like to further compare the genres that we found were most similar to determine what characteristics are the same and what characteristics are different. In that comparison we would like to include data on other characteristics that we did not include here. 

Another thing that we would be interesting in studying given more time is looking at other genres as well as including a few subgenres. We recognize that the genres that we studied are not all encompassing and there are other genres that would be interesting to collect data on. 

The final additional analysis that we would add would be to determine if there is a relationship between the characteristics that are similar between genres. For example we found that energy and loudness were similar for pop and latin which leads us to question whether there would be a relationship between the energy and the loudness of the track.

Ultimately while we can pick apart the different genres by looking at various characteristics of the tracks music is not something that picks us apart as people; music brings us together. Whether you like rap, enjoy some good R&B, or get down with some rock, music is what brings people from every walk of life together. So don't over think it just go pull up Spotify and relax to your favorite tunes.


# References:
(n.d.). Retrieved from R Wrapper for the 'Spotify' Web API: https://www.rcharlie.com/spotifyr/
Mock, T. (2020, January 21). Retrieved from TidyTuesday: https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-01-21)
Spotify. (2020, November 10). Retrieved from Wikipedia: https://en.wikipedia.org/wiki/Spotify

# Appendix: Summary Statistics and Results Tables

## Summary Statistics

```{r, warning=F, message=F, echo=F}
summary_table1<- group_by(spotify_songs, playlist_genre) %>% summarise(count=n(), mean_loudness=mean(loudness), mean_duration_in_ms=mean(duration_ms), mean_danceability=mean(danceability))

summary_table2<- group_by(spotify_songs, playlist_genre) %>% summarise(count=n(), mean_speechiness=mean(speechiness), mean_energry=mean(energy), mean_mean_popularity=mean(track_popularity))

summary_table1 %>% kable()
summary_table2 %>% kable()
```

```{r}
speechiness_aov <- aov(speechiness ~ playlist_genre, data=spotify_songs)
energy_aov <- aov(energy ~ playlist_genre, data=spotify_songs)
danceability_aov <- aov(danceability ~ playlist_genre, data=spotify_songs)
loudness_aov <- aov(loudness ~ playlist_genre, data=spotify_songs)
duration_aov <- aov(duration_ms ~ playlist_genre, data=spotify_songs)
popularity_aov <- aov(track_popularity ~ playlist_genre, data=spotify_songs)
```

## ANOVA for Speechiness

```{r}
tidy(speechiness_aov) %>% kable()
```

## ANOVA for Energy

```{r}
tidy(energy_aov) %>% kable()
```

## ANOVA for Danceability

```{r}
tidy(danceability_aov) %>% kable()
```

## ANOVA for Loudness

```{r}
tidy(loudness_aov) %>% kable()
```

## ANOVA for Duration

```{r}
tidy(duration_aov) %>% kable()
```

## ANOVA for Popularity

```{r}
tidy(popularity_aov) %>% kable
```

## Tukey Test for Speechiness
```{r}
tidy(TukeyHSD(speechiness_aov)) %>% kable()
```

## Tukey Test for Danceability

```{r}
tidy(TukeyHSD(danceability_aov)) %>% kable()
```

## Tukey Test for Energy

```{r}
tidy(TukeyHSD(energy_aov)) %>% kable()
```

## Tukey Test for Loudness

```{r}
tidy(TukeyHSD(loudness_aov)) %>% kable()
```

## Tukey Test for Duration

```{r}
tidy(TukeyHSD(duration_aov)) %>% kable()
```

## Tukey Test for Popularity

```{r}
tidy(TukeyHSD(popularity_aov)) %>% kable()
```

## Regression Output: Danceability

```{r}
danceability_model <- lm(data=spotify_songs, track_popularity~danceability)
get_regression_table(danceability_model) %>% kable()
get_regression_summaries(danceability_model) %>% kable()
```

## Regression Output: Energy

```{r}
energy_model <- lm(data=spotify_songs, track_popularity~energy)
get_regression_table(energy_model) %>% kable()
get_regression_summaries(energy_model) %>% kable()
```

## Regression Output: Speechiness

```{r}
speechiness_model <- lm(data=spotify_songs, track_popularity~speechiness)
get_regression_table(speechiness_model) %>% kable()
get_regression_summaries(speechiness_model) %>% kable()
```

## Regression Output: Duration

```{r}
duration_model <- lm(data=spotify_songs, track_popularity~duration_ms)
get_regression_table(duration_model) %>% kable()
get_regression_summaries(duration_model) %>% kable()
```

## Regression Output: Loudness

```{r}
loudness_model <- lm(data=spotify_songs, track_popularity~loudness)
get_regression_table(loudness_model) %>% kable()
get_regression_summaries(loudness_model) %>% kable()
```
