---
title: "Stat 850 Project Description"
author: "Neetu Regmi, Cassie Tangen, Oliver Titus ..."
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 3)
```

```{r load-sas-libraries, echo = F, message = F, warning = F}

saspath <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en" # Your SAS path goes here
sasopts <- "-ls 75"
knitr::opts_chunk$set(engine.path=list(sas=saspath, saslog=saspath, sashtml=saspath),
                      engine.opts=list(sas=sasopts, saslog=sasopts, sashtml=sasopts), 
                      comment=NA)
library(SASmarkdown)
```

## Instructions

Each member of your team should modify this document in some way and push their modifications to the repository in a separate commit. This will ensure that you have set your repository up in a way that ensures all group members are working with the same repository. 

Note that you must compile your readme (this document) for it to properly display as part of your github repository. 

Once you have received feedback on your project proposal (via Canvas) you may alter this README so that it describes your final project instead of the project proposal. 

## Data Set
```{r}
spotify_songs <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')
```


Data from TidyTuesday about songs on Spotify.
There are many variables included which are all described by TidyTuesday in the 01-21-2020 folder titled song genres.


## Potential Topics to Explore Using the Data Set
Describe avenues you might explore using the data

Cassie:
Exploring if there is difference in song length based on genre.
```{r}
library(tidyverse)
ggplot(spotify_songs, aes(x=duration_ms, fill=playlist_genre)) + geom_boxplot()
```

I decided to use a stacked boxplot to look at and compare the duration of the song duration in milliseconds of each different genre. It appears that rock and R&B have the longest durations and EDM is the shortest.

Create a few data frames that have been filtered for each genre. This will aid in calculations later.
```{r}
spotify_songs_edm<-filter(spotify_songs, playlist_genre=="edm")
spotify_songs_latin<-filter(spotify_songs, playlist_genre=="latin")
spotify_songs_pop<-filter(spotify_songs, playlist_genre=="pop")
spotify_songs_rb<-filter(spotify_songs, playlist_genre=="r&b")
spotify_songs_rap<-filter(spotify_songs, playlist_genre=="rap")
spotify_songs_rock<-filter(spotify_songs, playlist_genre=="rock")

```

Find out how tracks there are for each genre. This may be useful later.
```{r}
count_edm<-as.double(count(spotify_songs_edm))
count_latin<-as.double(count(spotify_songs_latin))
count_pop<-as.double(count(spotify_songs_pop))
count_rb<-as.double(count(spotify_songs_rb))
count_rap<-as.double(count(spotify_songs_rap))
count_rock<-as.double(count(spotify_songs_rock))


count<-c(count_edm,count_latin,count_pop,count_rb,count_rap,count_rock)

```


Let's look closer by looking numerically at the means. I'm also curious about how many songs there are for each genre so I am going to find the number of songs in each genre now as well. I will put the data into a table called genre_table
```{r}
# find mean duration in ms then change to min by /1.667*10^{-5}

mean_duration_edm<-mean(spotify_songs_edm$duration_ms)
mean_duration_latin<-mean(spotify_songs_latin$duration_ms)
mean_duration_pop<-mean(spotify_songs_pop$duration_ms)
mean_duration_rb<-mean(spotify_songs_rb$duration_ms)
mean_duration_rap<-mean(spotify_songs_rap$duration_ms)
mean_duration_rock<-mean(spotify_songs_rock$duration_ms)

mean_duration<-c(mean_duration_edm/1.667*10^{-5},mean_duration_latin/1.667*10^{-5},mean_duration_pop/1.667*10^{-5},mean_duration_rb/1.667*10^{-5},mean_duration_rap/1.667*10^{-5},mean_duration_rock/1.667*10^{-5})

genre<-c("EDM","Latin","Pop","R&B","Rap","Rock")

genre_table<-tibble(genre,mean_duration)
print(genre_table)

```

It appears that the genre with the longest songs is rock and the genre with the shortest songs is rap. Would be interesting to perform a hypothesis test to see if there is a statitical difference between the lengths of songs.


Explore the popularity of each genre.
```{r}
ggplot(spotify_songs, aes(x=track_popularity, fill=playlist_genre)) + geom_boxplot()
```
It appears that pop and latin on average are more popular and EDM is the least popular.
Let's go ahead and calculate mean popularity and add it to the genre table

```{r}
mean_popularity_edm<-mean(spotify_songs_edm$track_popularity)
mean_popularity_latin<-mean(spotify_songs_latin$track_popularity)
mean_popularity_pop<-mean(spotify_songs_pop$track_popularity)
mean_popularity_rb<-mean(spotify_songs_rb$track_popularity)
mean_popularity_rap<-mean(spotify_songs_rap$track_popularity)
mean_popularity_rock<-mean(spotify_songs_rock$track_popularity)

mean_popularity<-c(mean_popularity_edm,mean_popularity_latin,mean_popularity_pop,mean_popularity_rb,mean_popularity_rap,mean_popularity_rock)


genre_table<-tibble(genre,mean_duration)%>% mutate(mean_popularity)%>%mutate(count)
print(genre_table)
```
It appears that EDM is the least popular on average, but it has the most songs so that could be impacting the popularity. I would be interested in performing a hypothesis test to see if EDM's mean popularity is statistically different than that of the others that all appear to be roughly the same popularity-wise.


Explore the loudness of different genre.
```{r}
ggplot(spotify_songs, aes(x=loudness, fill=playlist_genre)) + geom_boxplot()
```
EDM appears to be the loudest with R&B the least loud. Let's look at the means numerically as well.
```{r}
mean_loud_edm<-mean(spotify_songs_edm$loudness)
mean_loud_latin<-mean(spotify_songs_latin$loudness)
mean_loud_pop<-mean(spotify_songs_pop$loudness)
mean_loud_rb<-mean(spotify_songs_rb$loudness)
mean_loud_rap<-mean(spotify_songs_rap$loudness)
mean_loud_rock<-mean(spotify_songs_rock$loudness)

mean_loudness<-c(mean_loud_edm,mean_loud_latin,mean_loud_pop,mean_loud_rb,mean_loud_rap,mean_loud_rock)

genre_table<-tibble(genre,mean_duration)%>% mutate(mean_popularity)%>%mutate(mean_loudness)%>%mutate(count)
print(genre_table)
```
It seems R&B is the least loud and EDM is the loudest. There is not a huge spread between genre's for loudness, but it does seem that EDM is a decent amount louder than the other varieties. I would be interesting in running a hypothesis test here as well.




Explore track artist of different genre.

Explore a potential relationship between tempo and song duration.

Explore a potential relationship between tempo and dancibility,



Explore the dancibility, energy, key,mode, speechiness, etc based on genre.

Explore the relationship between dancibility and loudness.

Explore the relationship between dancibility and temp.


## Group Members
List all of the project contributors here.
Neetu Regmi 
Oliver Titus 
Cassie Tangen
